---
title: | 
  | Data analysis report - 
  | The effect of an exercise intervention program on bone health after bariatric surgery: BaSEIB clinical trial
author: "Lucas Veras"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: 
      collapsed: true
    code_folding: "show"
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This report covers the content of a [github repository](https://github.com/verasls/BaSEIB_exercise_bone_health) that encompasses the data analysis of a manuscript entitled "The effect of an exercise intervention program on bone health after bariatric surgery: BaSEIB clinical trial". The [code](https://github.com/verasls/BaSEIB_exercise_bone_health/tree/master/code) folder contains two subfolders: [functions](https://github.com/verasls/BaSEIB_exercise_bone_health/tree/master/code/functions), with some function definitions, and [scripts](https://github.com/verasls/BaSEIB_exercise_bone_health/tree/master/code/scripts), with 9 different scripts regarding distinct steps of the data processing and statistical analysis. Also, code to generate the manuscript figures are inside the [figs](https://github.com/verasls/BaSEIB_exercise_bone_health/tree/master/figs) folder.

# Load and tidy data

The [01_tidy_data.R](https://github.com/verasls/BaSEIB_exercise_bone_health/blob/master/code/scripts/01_tidy_data.R) script loads the database.csv file. It, then, selects, renames and recodes some variables and excludes data from subjects who do not have at least one follow-up assessment. A part of the resulting data frame, that is the base for all subsequent analyses, can be seen below.

```{r loads, echo=FALSE}
library(here)
source(here("code", "scripts", "01_tidy_data.R"))
```


```{r df, rows.print=12}
df
```